\AorE{Wavelet Scattering Transformasies Toegepas op Walvis Vokaliserings}{Wavelet Scattering Transforms Applied to Whale Vocalisations}

### 2024-05-27 dissertation/content/paper2/main.tex
started intro to p3
added astract and conclusion
draft edit nearly complete, need to do refs

\chapter{Introducing Wavelet Scattering}

\articleinfo{M.W. Rademan, D.J.J Verseld, J.A. Du Preez}{the Journal of the Acoustical Society of America (JASA)}{15 May 2024}{doi}{Current literature trends for whale classification in passive acoustic monitoring systems are mainly concerned with high accuracies obtained via deep learning approaches. This requires preparation of large and accurate annotated datasets, which are often not available when surveying new regions, nor are they wholly accurate when they are. In contrast to classification, many efforts are still being made in improving generalised detectors for the purpose of disregarding noise. As such, there is a disparity between detectors and classification models.

@@

The proposed method works well for small datasets and aims to form the foundation of wavelet scattering classification/detection using non-neural network approaches. The purpose of this work is to provide a practical classification and detection system for new regions in which very little annotated data is available.  }

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\textwidth]{system_diagram_v2.png}
    \caption{The detection and classifying system. The audio database constists of 1-hour long recordings, which are segmented for computational efficiency. The detection system provides input to the feature extraction and classification steps. The classifier and detector are evaluated separately in this study.}
    \label{fig:system}
\end{figure}

@@

\begin{figure}
     \centering
     \begin{subfigure}[]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{Bm_Ant-A_11.jpg}
     \end{subfigure}
     \hfill
     \begin{subfigure}[]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{Bm_Ant-B_54.jpg}
     \end{subfigure}
     \hfill
     \begin{subfigure}[]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{Bm_Ant-Z_98.jpg}
     \end{subfigure}
     \hfill
     \begin{subfigure}[]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{Bm_D_173.jpg}
     \end{subfigure}
        \caption{High SNR examples of the first level scattering transform to illustrate the characteristics of blue whale calls. Note the logarithmic frequency scale.  \textit{Bm\_Ant\_A} (a) is a constant frequency pulse which occurs around 20 Hz. \textit{Bm\_Ant\_B} (b) and \textit{Bm\_Ant\_Z} (c) are similar to \textit{Bm\_Ant\_A}, but with the addition of frequency modulated tails. \textit{Bm\_D} (d) is a frequency modulated downsweep which typically starts around the 80 Hz and sweeps to 30 Hz.}
        \label{fig:Bm_Ex}
\end{figure}

@@

\begin{table}[h!]
    \centering
\caption{Estimated erroneous (false positives) proportion of annotations per class that could be considered noise. This table does not reflect on whether annotations are of the correct class (i.e., misclassifications). The total proportion of false positives are calculated according to the number of annotations per class.}
\label{tab:errannotations}
% \begin{ruledtabular}    
\begin{tabular}{lr} 
\textbf{Vocalisation}      & \textbf{False positives (\%)} \\ \hline
Bm\_Ant\_A & 5 - 10                 \\
Bm\_Ant\_B & 5 - 10                  \\
Bm\_Ant\_Z & 1 - 2                  \\
Bm\_D      & 8 - 15                  \\
Bp\_20Hz & 1 - 3                   \\
Bp\_20Plus & 1 - 3                   \\
Total & 5 - 10\\ 
\end{tabular}
% \end{ruledtabular}
\end{table}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\textwidth]{scattering.png}
    \caption{An illustration of the successive filtering operations performed by a scattering transform. Nodes in colour represent the extracted scattering coefficients, which is down-sampled according to an invariance scale specified by the user.}
    \label{fig:scattering}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=1\textwidth]{fb.jpg}
    \caption{The filter banks used in this study, with $Q_1 = 16$ and $Q_2 = 4$. The filter bank used to calculate $\mathcal{U}_1$ is shown on the left and the $\mathcal{U}_2$ filter bank on the right. Note that the maximum magnitude of 2 accounts for the analytic nature of the wavelet, thereby capturing the true frequency amplitudes in a signal.}
    \label{fig:fb}
\end{figure}

\clearpage
\section{Improved Spectral Entropy Detector}
The \ac{se} detector described by \citet{mypaper} is improved in this section. Rademan shows empirically that the \ac{cwt} yields improved results compared to the \ac{stft} for practical \acp{snr}. As such, we utilise the first level scattering coefficients $\mathcal{S}_{1}$, since it is equivalent to a down sampled version of the \ac{cwt}. We improve on the detector posed by Rademan by introducing true probabilistic output and spectral noise whitening.

@@

\begin{figure}[]
    \centering
    \includegraphics[width=1\textwidth]{detector_example_2.eps}
    \caption{An example of the output of the detector taken from a densely populated vocalisation region of the dataset. The annotation bounding boxes are also indicated (top). In this example, the detector successfully detects all vocalisations. The whitened SE measure is shown (middle) with the posterior probability of signal (bottom).}
    \label{fig:det_example_1}
\end{figure}

@@

\begin{figure}[h]
     \centering
     \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \includegraphics[width=1\textwidth]{roc.eps}    
     \caption{} \label{fig:det_roc}
     \end{subfigure}
     % \hfill
     \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \includegraphics[width=1\textwidth]{calls.eps}    
    \caption{} \label{fig:det_calls}
     \end{subfigure}
     \caption{Results of the evaluated detectors. \ac{se} detectors are indicated as a solid line, while \ac{bled} is dashed. (a) The performance of the detectors when compared on a sample-by-sample bases. (b) The performance curve of the detectors expressed as the percentage of detected noise samples versus the percentage of detected calls. The whitened \ac{se} with \ac{gmm} detector significantly outperforms all other detectors.}
\end{figure}

@@

\clearpage
\section{Classification}
\label{sec:cls}
We combine the output of the \ac{se} detector with \ac{ws} features and utilise \ac{lda} to classify the detected calls. Our method shares many similarities with studies that also employ \ac{ws} without a \ac{nn}, but differs in allowing for variable window sizes produced by a detector \citep{ws_audio, ws_ecg}.

@@

\begin{table}[h!]
\caption{Number of labeled detections for all vocalisations of interest. The detections are produced by the whitened \ac{gmm} \ac{se} detector. A significant number of detections are discarded in the training data due to overlapping.}
\label{tab:numdetections}
% \begin{ruledtabular}
% \renewcommand{\arraystretch}{0.8}
\begin{tabular}{lc} 
\textbf{Vocalisation}      & \textbf{Number of detections}  \\ \hline
Bm\_Ant\_A & 1514              \\
Bm\_Ant\_B & 474              \\
Bm\_Ant\_Z & 91               \\
Bm\_D      & 328               \\
Bp & 50                \\
Noise      & 25638          \\
Overlapping & 150 \footnotemark[1]      \\ 
\end{tabular}
% \end{ruledtabular}
\footnotetext[1]{Overlapping detections contain at least 2 overlapping annotations of different classes. This implies that more than 300 vocalisations are discarded as a result of overlap, thereby drastically reducing training data for the \textit{Bm\_D} and \textit{Bp\_20Plus} classes, which have a higher tendency for overlap.}
\end{table}

@@

\begin{table}[]
\caption{Average model classification accuracies for varying train-test splits with chi-squared feature selection. An 80\% training split without chi-square feature selection is included as a baseline. 20 trials are performed, whereas 5\% splits use 100 trials. The median accuracy is displayed.}
\label{tab:clsres}
% \begin{ruledtabular}  
\renewcommand{\arraystretch}{0.8}
\begin{tabular}{lcccc}
\textbf{Vocalisation} & \textbf{Vocalisation Accuracy (\%)} & \textbf{Noise Accuracy (\%)} & $\alpha$ & $\gamma$  \\ \hline  
\multicolumn{5}{c}{80\% Train (No Feature Selection)} \\ \hline 
Bm\_Ant\_A & 82.2 ± 1.5  & 88.6 ± 1.2 & 0.66 & 0.71 \\
Bm\_Ant\_B & 83.3 ± 6.2  & 92.3 ± 1.1 & 0.54 & 0.43 \\
Bm\_Ant\_Z & 92.5 ± 10.6 & 98.2 ± 0.1 & 0.05 & 0.43 \\
Bm\_D      & 74.1 ± 6.6  & 90.3 ± 2.0 & 0.66 & 0.86 \\
Bp         & 96.0 ± 14.6 & 99.0 ± 0.1 & 0.17 & 0.57\\  \hline

@@

\clearpage
\section{Conclusion}

@@

    \begin{table}[h]
        \caption{Summary of parameter values used in this study. Chosen values are indicated in boldface text, whereas non-bold parameters result from the chosen parameters.}
        \label{tab:parameters}
        % \begin{ruledtabular}    
            \renewcommand{\arraystretch}{0.8}% Tighter
            \begin{tabular}{lll}  \small
                \textbf{Parameter}      & \textbf{Description} & \textbf{Value} \\ \hline
                $T$ & \ac{ws} invariance scale & \textbf{1.5 seconds} \\
                $Q_1$ & Wavelets per octave in $\mathcal{S}_1$ & \textbf{16} \\
                $Q_2$ & Wavelets per octave in $\mathcal{S}_2$ & \textbf{4} \\
                $f_\text{min}$ & Minimum filterbank frequency in $\mathcal{S}_1$ & \textbf{15 Hz}\\
                $f_\text{max}$ & Maximum filterbank frequency in $\mathcal{S}_1$ & \textbf{120 Hz}\\
                $|\Lambda_1|$ & Resulting number of filters in $\mathcal{S}_1$ & 48 \\
                $|\Lambda_2|$ & Resulting number of filters in $\mathcal{S}_2$ & 14 \\
                $V$ & Wavelet scattering upsample factor & \textbf{2} \\
                $r$ & Downsampling factor of $\mathcal{U}_1$ due to maximum bandwidth & 8 \\
                $d$ & Downsampling factor of $\mathcal{S}_1$ & 5 \\
                $f_{ss}$ & Resulting scattering transform sample frequency & 6.25 Hz \\
                $B$ & Median filter length for noise coefficient suppression & \textbf{9} \\
                $D$ & Median filter length for transient coefficient suppression  & \textbf{5 seconds} \\
                $F$ & Average filter length for noise coefficient estimation  & \textbf{60 seconds} \\
                $R$ & Median filter length for stabilizing entropy  & \textbf{1.5 seconds} \\
                $G$ & Maximum filter length for boundary extension  & \textbf{6 seconds} \\
                $p_t$ & Probability threshold for the whitened \ac{se} \ac{gmm} detector & \textbf{0.3} \\ 
                $k$ & Dimensions in the classification feature vector $v$ & 521\\ 
                \end{tabular}
                % \end{ruledtabular}
                \end{table}

>>>>

\chapter{Introducing Wavelet Scattering}
\label{chap:p2}

\articleinfo{Detecting and classifying Whale Calls with Wavelet Scattering
and Spectral Entropy}{the Journal of the Acoustical Society of America (JASA)}{15 May 2024}{doi}

\section*{CRediT}
\textbf{M.W. Rademan}: conceptualisation; methodology; software; formal analysis; writing -- original draft.

\textbf{D.J.J. Verseld}: writing -- review and editing; supervision

\textbf{J.A. Du Preez}: writing -- review and editing; supervision.

\section*{Abstract}
Current literature trends for whale classification in passive acoustic monitoring systems are mainly concerned with high accuracies obtained via deep learning approaches. This requires preparation of large and accurate annotated datasets, which are often not available when surveying new regions, nor are they wholly accurate when they are. In contrast to classification, many efforts are still being made in improving generalised detectors for the purpose of disregarding noise. As such, there is a disparity between detectors and classification models.

@@

The proposed method works well for small datasets and aims to form the foundation of wavelet scattering classification/detection using non-neural network approaches. The purpose of this work is to provide a practical classification and detection system for new regions in which very little annotated data is available.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\textwidth]{system_diagram_v2.png}
    \caption[The detection and classifying system.]{The detection and classifying system. The audio database constists of 1-hour long recordings, which are segmented for computational efficiency. The detection system provides input to the feature extraction and classification steps. The classifier and detector are evaluated separately in this study.}
    \label{fig:system}
\end{figure}

@@

\begin{figure}
     \centering
     \begin{subfigure}[]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{Bm_Ant-A_11.jpg}
     \end{subfigure}
     \hfill
     \begin{subfigure}[]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{Bm_Ant-B_54.jpg}
     \end{subfigure}
     \hfill
     \begin{subfigure}[]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{Bm_Ant-Z_98.jpg}
     \end{subfigure}
     \hfill
     \begin{subfigure}[]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{Bm_D_173.jpg}
     \end{subfigure}
        \caption[High SNR examples of the first level scattering transform to illustrate the characteristics of blue whale calls.]{High SNR examples of the first level scattering transform to illustrate the characteristics of blue whale calls. Note the logarithmic frequency scale.  \textit{Bm\_Ant\_A} (a) is a constant frequency pulse which occurs around 20 Hz. \textit{Bm\_Ant\_B} (b) and \textit{Bm\_Ant\_Z} (c) are similar to \textit{Bm\_Ant\_A}, but with the addition of frequency modulated tails. \textit{Bm\_D} (d) is a frequency modulated downsweep which typically starts around the 80 Hz and sweeps to 30 Hz.}
        \label{fig:Bm_Ex}
\end{figure}

@@

\begin{table}[h!]
    \centering
\caption[Estimated erroneous (false positives) proportion of annotations per class that could be considered noise.]{Estimated erroneous (false positives) proportion of annotations per class that could be considered noise. This table does not reflect on whether annotations are of the correct class (i.e., misclassifications). The total proportion of false positives are calculated according to the number of annotations per class.}
\label{tab:errannotations}
% \begin{ruledtabular}    
\begin{tabular}{lr} 
\textbf{Vocalisation}      & \textbf{False positives (\%)} \\ \hline
Bm\_Ant\_A & 5 - 10                 \\
Bm\_Ant\_B & 5 - 10                  \\
Bm\_Ant\_Z & 1 - 2                  \\
Bm\_D      & 8 - 15                  \\
Bp\_20Hz & 1 - 3                   \\
Bp\_20Plus & 1 - 3                   \\
Total & 5 - 10\\ 
\end{tabular}
% \end{ruledtabular}
\end{table}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\textwidth]{scattering.png}
    \caption[An illustration of the successive filtering operations performed by a scattering transform.]{An illustration of the successive filtering operations performed by a scattering transform. Nodes in colour represent the extracted scattering coefficients, which is down-sampled according to an invariance scale specified by the user.}
    \label{fig:scattering}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=1\textwidth]{fb.jpg}
    \caption[The filter banks used in this study.]{The filter banks used in this study, with $Q_1 = 16$ and $Q_2 = 4$. The filter bank used to calculate $\mathcal{U}_1$ is shown on the left and the $\mathcal{U}_2$ filter bank on the right. Note that the maximum magnitude of 2 accounts for the analytic nature of the wavelet, thereby capturing the true frequency amplitudes in a signal.}
    \label{fig:fb}
\end{figure}

\section{Improved Spectral Entropy Detector}
The \ac{se} detector described by \citet{mypaper} is improved in this section. Rademan shows empirically that the \ac{cwt} yields improved results compared to the \ac{stft} for practical \acp{snr}. As such, we utilise the first level scattering coefficients $\mathcal{S}_{1}$, since it is equivalent to a down sampled version of the \ac{cwt}. We improve on the detector posed by Rademan by introducing true probabilistic output and spectral noise whitening.

@@

\begin{figure}[]
    \centering
    \includegraphics[width=1\textwidth]{detector_example_2.eps}
    \caption[An example of the output of the detector.]{An example of the output of the detector taken from a densely populated vocalisation region of the dataset. The annotation bounding boxes are also indicated (top). In this example, the detector successfully detects all vocalisations. The whitened SE measure is shown (middle) with the posterior probability of signal (bottom).}
    \label{fig:det_example_1}
\end{figure}

@@

\begin{figure}[h]
     \centering
     \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \includegraphics[width=1\textwidth]{roc.eps}    
     \caption{} \label{fig:det_roc}
     \end{subfigure}
     % \hfill
     \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \includegraphics[width=1\textwidth]{calls.eps}    
    \caption{} \label{fig:det_calls}
     \end{subfigure}
     \caption[ROC curves of the evaluated detectors.]{Results of the evaluated detectors. \ac{se} detectors are indicated as a solid line, while \ac{bled} is dashed. (a) The performance of the detectors when compared on a sample-by-sample bases. (b) The performance curve of the detectors expressed as the percentage of detected noise samples versus the percentage of detected calls. The whitened \ac{se} with \ac{gmm} detector significantly outperforms all other detectors.}
\end{figure}

@@

\section{Classification}
\label{sec:cls}
We combine the output of the \ac{se} detector with \ac{ws} features and utilise \ac{lda} to classify the detected calls. Our method shares many similarities with studies that also employ \ac{ws} without a \ac{nn}, but differs in allowing for variable window sizes produced by a detector \citep{ws_audio, ws_ecg}.

@@

\begin{table}[h!]
\caption[Number of labeled detections for all vocalisations of interest.]{Number of labeled detections for all vocalisations of interest. The detections are produced by the whitened \ac{gmm} \ac{se} detector. A significant number of detections are discarded in the training data due to overlapping.}
\label{tab:numdetections}
% \begin{ruledtabular}
% \renewcommand{\arraystretch}{0.8}
\centering
\begin{tabular}{lc} 
\textbf{Vocalisation}      & \textbf{Number of detections}  \\ \hline
Bm\_Ant\_A & 1514              \\
Bm\_Ant\_B & 474              \\
Bm\_Ant\_Z & 91               \\
Bm\_D      & 328               \\
Bp & 50                \\
Noise      & 25638          \\
Overlapping & 150 \footnotemark[1]      \\ 
\end{tabular}
% \end{ruledtabular}
\footnotetext[1]{Overlapping detections contain at least 2 overlapping annotations of different classes. This implies that more than 300 vocalisations are discarded as a result of overlap, thereby drastically reducing training data for the \textit{Bm\_D} and \textit{Bp\_20Plus} classes, which have a higher tendency for overlap.}
\end{table}

@@

\begin{table}[]
\caption[Average model classification accuracies for varying train-test splits with chi-squared feature selection.]{Average model classification accuracies for varying train-test splits with chi-squared feature selection. An 80\% training split without chi-square feature selection is included as a baseline. 20 trials are performed, whereas 5\% splits use 100 trials. The median accuracy is displayed.}
\label{tab:clsres}
% \begin{ruledtabular}  
\renewcommand{\arraystretch}{0.8}
\begin{tabular}{lcccc}
\textbf{Vocalisation} & \textbf{Vocalisation Accuracy (\%)} & \textbf{Noise Accuracy (\%)} & $\alpha$ & $\gamma$  \\ \hline  
\multicolumn{5}{c}{80\% Train (No Feature Selection)} \\ \hline 
Bm\_Ant\_A & 82.2 ± 1.5  & 88.6 ± 1.2 & 0.66 & 0.71 \\
Bm\_Ant\_B & 83.3 ± 6.2  & 92.3 ± 1.1 & 0.54 & 0.43 \\
Bm\_Ant\_Z & 92.5 ± 10.6 & 98.2 ± 0.1 & 0.05 & 0.43 \\
Bm\_D      & 74.1 ± 6.6  & 90.3 ± 2.0 & 0.66 & 0.86 \\
Bp         & 96.0 ± 14.6 & 99.0 ± 0.1 & 0.17 & 0.57\\  \hline

@@

\section{Conclusion}

@@

    \begin{table}[h]
        \caption[Summary of parameter values used in this study. ]{Summary of parameter values used in this study. Chosen values are indicated in boldface text, whereas non-bold parameters result from the chosen parameters.}
        \label{tab:parameters}
        % \begin{ruledtabular}    
            % \renewcommand{\arraystretch}{0.8}% Tighter
            \begin{tabular}{lll} 
                \textbf{Parameter}      & \textbf{Description} & \textbf{Value} \\ \hline
                $T$ & \ac{ws} invariance scale & \textbf{1.5 seconds} \\
                $Q_1$ & Wavelets per octave in $\mathcal{S}_1$ & \textbf{16} \\
                $Q_2$ & Wavelets per octave in $\mathcal{S}_2$ & \textbf{4} \\
                $f_\text{min}$ & Minimum filterbank frequency in $\mathcal{S}_1$ & \textbf{15 Hz}\\
                $f_\text{max}$ & Maximum filterbank frequency in $\mathcal{S}_1$ & \textbf{120 Hz}\\
                $|\Lambda_1|$ & Resulting number of filters in $\mathcal{S}_1$ & 48 \\
                $|\Lambda_2|$ & Resulting number of filters in $\mathcal{S}_2$ & 14 \\
                $V$ & Wavelet scattering upsample factor & \textbf{2} \\
                $r$ & Downsampling factor of $\mathcal{U}_1$ due to maximum bandwidth & 8 \\
                $d$ & Downsampling factor of $\mathcal{S}_1$ & 5 \\
                $f_{ss}$ & Resulting scattering transform sample frequency & 6.25 Hz \\
                $B$ & Median filter length for noise coefficient suppression & \textbf{9} \\
                $D$ & Median filter length for transient coefficient suppression  & \textbf{5 seconds} \\
                $F$ & Average filter length for noise coefficient estimation  & \textbf{60 seconds} \\
                $R$ & Median filter length for stabilizing entropy  & \textbf{1.5 seconds} \\
                $G$ & Maximum filter length for boundary extension  & \textbf{6 seconds} \\
                $p_t$ & Probability threshold for the whitened \ac{se} \ac{gmm} detector & \textbf{0.3} \\ 
                $k$ & Dimensions in the classification feature vector $v$ & 521\\ 
                \end{tabular}
                % \end{ruledtabular}
                \end{table}
